Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-03-25 paper/GEB Submission/manuscript_GEB.tex
Mooore edits

\begin{abstract}
Aim: Despite unprecedented environmental change due to anthropogenic
pressure, recent work has found increasing species turnover but no
overall trend in species diversity through time at the local scale.
Functional diversity provides a potentially powerful alternative
approach for understanding community composition by linking shifts in
species identity to mechanisms of ecosystem processes. Here we present
the first multi-taxa, multi-system analysis of functional change through
time.

@@

Results: We found no overall trend in any functional diversity metric,
despite data replicating species-based patterns of constant richness
with increasing turnover. The lack of trend held even after correcting
for changes in species richness. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however most studies showed a shift in a species or functional
metric.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, we identified four
prevailing scenarios of species and functional change, which showed
links to the duration of the observation window. With no one prevailing
scenario of change, it will be critical to link change scenarios to
drivers of change, particularly to identify communities with capacity to
resist drivers from those not experiencing substantial pressure from a
driver.
\end{abstract}

@@

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and exclusion of individual
studies' ecological context (10--12). Many of these criticisms reflect
limitations of ecological data on the whole,
\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
(13, 14).

@@

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While functional loss is frequently cited as one of the most
pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
functional composition shows mixed responses to environmental change and
extinction events (30, 31), with significant impacts of species
extinctions on functional diversity in some taxa and not others (32).
For some time periods, functional structure appears to be maintained for
substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
(34).

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as a
significant subset of the world's biodiversity heavily impacted by
anthropogenic change. While examining trends in plants, invertebrates,
and other vertebrate species is of equal interest, trait data for those
taxa raise additional challenges such as limited and biased species
coverage (35), a lack of accepted species-level means, and differences
in the types of traits collected. To ensure comparability across taxa in
trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(31). All time series include abundance of observed species. Following
best practices for the database (37), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

@@

Our final dataset included 2,432 time series from 50 studies in 21
countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five climates
(Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
earliest sample was in 1923 and the most recent was in 2014.
\textcolor{blue}{While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level. However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.}
For a full breakdown of studies and their characteristics, see the
supplement. Our final dataset reflects many of the data biases that make
global synthesis work challenging, including geographic bias, a bias
away from areas currently under the greatest threat, and a bias towards
shorter time series. We address these shortcomings and their potential
impact on our results in the discussion.

@@

We calculated yearly metrics of functional and species diversity for
each time series. Species-based metrics include species richness
(\emph{S}) and Jaccard similarity (\emph{J}) as a measure of turnover.
Jaccard similarity was calculated relative to the first observed year
for a time series. A negative trend in \emph{J} would therefore indicate
increasing turnover. We did not impose a correction for unobserved
species as non-parametric estimators do not assign species identities to
corrected richness values, and therefore could not be propagated to the
functional diversity metrics.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (46). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species. `
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.

@@

We estimated general trends
\textcolor{blue}{across bird and mammals communities} for each diversity
metric using a linear mixed effects model with a random slope and
intercept for each study and each time series nested within the study.
We fit 18 individual \emph{CWM} models. All time series with data for a
given trait were included in the corresponding \emph{CWM} model. We
estimated study level trends using individual linear models. For studies
with more than one times series we fit a random slope and intercept for
time series. Some study-level models could not be fit for five studies
for at least one metric due to data limitations, but those studies were
still included in the general models. They represented 12 of 1350
study-level models fit for each metric. For further details see the
supplement. Where appropriate, response variables were \(log\) or
\(log(x+1)\) transformed to better fit model assumptions of residual
normality.

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 47). We assessed the
impact of time series duration and start year on study-level trends
using linear models with duration and start year as predictors. All
models in our analysis were fit using the \emph{lme4} (v1.1-30) package
in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of
freedom method using the \emph{lmerTest} (v3.1-3) package with a
significance level of \(\alpha = 0.05\) (48--50).

@@

We found no significant overall trend in species richness or functional
diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}).
\textcolor{blue}{Within-group} trends for different taxa, biomes,
realms, or protection statuses were also non-significant for richness
and functional diversity metrics, with the exception of a significantly
increasing trend for functional evenness of global studies
(characterized by having samples on multiple continents), and a
significantly decreasing functional richness slope for
temperate/tropical studies and mammal studies. However, trends were not
exhibited for the \textcolor{blue}{corrected} metric indicating that
differences \textcolor{blue}{in functional diversity metrics} were
largely due to changes in species richness. Further, with only two
global studies, the trend should not be considered truly general. The
general trends for \emph{CWM} models were similarly not significant,
with a significant positive trend for only percentage of fish in diet
composition.

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/3met_long} \caption{Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES}\label{fig:timeseriesPlot}
\end{figure}

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example, while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. We also found a significant decrease in Jaccard
similarity in unprotected areas only, with no trend for protected areas.
For bird communities, we also found within group trends and between
group differences for trends in foraging behavior. We found a
significant increasing trend in utilization of the canopy in Tropical
communities that was significantly different from the trends for
Polar/Temperate and Temperate communities. There was a significant
decrease in utilization of the understory for Terrestrial communities
and significant increase in foraging below the water surface for global
studies (but see the previous limitations of Global data). There was
also a significant positive slope for \emph{CWM} body mass for the
single Temperate/Tropical study (with three time series) which were
marine mammal communities.

We found significant dietary shifts across communities, with a
significant increase in fruit consumption in Terrestrial communities and
a significant decrease in nectar consumption in Tropical communities, a
trend significantly different than that for Terrestrial communities.
There was a significant increase in seed consumption in bird species,
which was significantly different from the trend for Mammal communities.
There was a significant increase in fish consumption for the two Global
studies. Vertebrate consumption significantly decreased for Marine
studies and for studies of Global and Tropical communities.
\textcolor{blue}{There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.}

At the study level, 15 of 50 studies exhibited a significant trend in
species richness and 11 exhibited significant turnover. For observed
functional metrics, 25 of 50 studies exhibited a trend in a least one
metric, and 15 of 50 studies exhibited a significant trend for at least
one \textcolor{blue}{corrected} metric (Table \ref{tab:trendTab}). In
general, there were more significant trends for uncorrected metrics,
with some disappearing after correction, indicating that those trends
were likely due to changes in the number of species. Hypothesis testing
for study-level trends is likely affected by multiple testing issues and
some trends identified as significant are therefore potentially
spurious. Rather than interpreting changes in specific studies, we
present these results as a general picture of the kinds of trends
experienced by communities.

Study-level slopes for functional diversity metrics were significantly
related to start year of the time series for Jaccard similarity and
functional evenness, both of which had significantly more negative
slopes with more recent start year. No functional diversity metrics were
significantly related to the start year of the time series.

We assessed the sensitivity of general trend results to major data
processing decision by rerunning models with increasingly conservative
subsets of the data. After excluding time series with less than two,
three, and four year durations, we found that the general trend for
increase in CWM aerial foraging disappeared, but two other trends in
CWM's appeared. After excluding time series two years long, a general
trend appeared for increased CWM use of the forest canopy, and decreased
CWM of seed consumption. These two trends remained as increasingly
shorter time series were excluded suggesting the increase in aerial
foraging was an erroneous finding while the other two trends were more
robust. Percentage of species in a community with trait data did not
appear to have an affect on general trend results, as general trends
were unchanged with increasing cut offs for percentage of species with
trait data. A complete list of models run in the sensitivity analysis
and their results can be found in the supplement.

@@

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
We did not detect an overall trend in any functional diversity metrics.
As with previous species-based syntheses, we also found no overall trend
in species richness accompanied by increasing turnover through time
(31), indicating that non-significant trends in functional metrics are
consistent with similar well-documented species derived trends. We found
no trend in functional change for almost all realms, biomes, and
taxonomic groups.
\textcolor{blue}{The one exception was a consensus trend indicating}
functional richness loss for mammal studies. These results are
consistent with multiple studies linking anthropogenic drivers to loss
of functional diversity in mammal communities (51--53).

Despite a lack of general trends in functional diversity metrics, we did
find evidence of compositional shifts in the trait space with multiple
\texttt{significant} trends in \emph{CWM}'s for studies coming from the
same taxa, realms, climates, or protection statuses,
\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
Some of those trends are consistent with existing work in those
contexts. For example, the decrease in insect consumption for tropical
communities reflects well documented declines in tropical insectivorous
birds (54). Reduction in utilization of understory foraging for
terrestrial communities could be the result of disturbance like human
recreation or increased predation from introduced species. Other shifts
in diet for both Birds and Mammals, including increasing fruit, seed,
and vertebrate consumption and decreasing seed consumption in some
climates point to important areas for further exploration.

@@

\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens. The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level. In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend. While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover. We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives. We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.}

@@

The lack of trend could be the result of multiple possible scenarios.
First, these may be communities resisting perturbations or simply not
experiencing significant perturbations. Given the studies in this group
come from all possible taxa, realms, climates, and protection statuses,
evidence points to communities resisting perturbation, offering some
hope that there are areas of the globe where communities are fairing
well for now. Alternatively, these may be communities that have
experienced or continue to experience significant stress, but lost
species or functional diversity outside the observation window. This
could be true particularly for North American mammal communities where
trophic downgrading and megafaunal losses occurred hundreds of years ago
{[}citation{]}. Third, these communities may be experiencing directional
shifts undetectable by available data. For example, species-level trait
data does not capture intraspecific shifts in the trait space, which can
represent significant changes in overall functional composition and
impact maintenance of ecological processes {[}citation{]}.

Three of the groups fit under the broad umbrella of changes in
redundancy. By definition, if a community gains or loses species while
functional metrics are unchanged, those species represent an increase or
decrease in redundancy, so we include communities exhibiting gains or
losses in species richness with no change in functional metrics or
turnover and richness loss with significant turnover in this umbrella.
For communities exhibiting loss of redundancy (declining species
richness), ecological processes are likely being maintained, but
capacity to respond to future stressors is reduced (55). These
communities are actually fairing better than expected looking at
species-based metrics alone, but are also in a precarious position for
maintaining ecological function into the future (56). Conversely,
communities exhibiting a increase in redundancy (via an increase in
species richness) are becoming better positioned to respond to future
stressors, but are not actually expanding their functional capacity as
may be assumed by looking at species gains alone (57).

The next group of studies exhibit an increase in species richness and
functional richness with significant turnover. While these communities
are losing some species that are functionally redundant or functional
analogs of species additions, they are gaining even more species that
expand the functional space. These communities have a hopeful
trajectory, as they are improving their functional capacity and
potentially their functional redundancy, likely leading to robust future
ecological function. Notably, the studies that exhibited species
richness increases (in this group with increasing functional richness,
as well as communities without functional richness increases) were
exclusively from terrestrial, temperate (sometimes temperate/polar) bird
communities and occurred in countries where there has been a significant
investment in conservation over the last few decades (United States,
Canada, Sweden). Our results are consistent with other functional work
in these regions, showing increases in species richness and functional
diversity for North American breeding birds over the last few decades
after a period of decline(58), and loss of common, functionally general
species even as rare species are increasing in North America and Europe
(59--61).

The final group exhibits decreasing functional evenness. The empirical
link between changes in evenness and ecosystem process is the most
poorly studied of the metrics, however theory indicates that increased
evenness improves function and stability (62). If one or a few species
become largely dominant in a community, function will be the result of
only those species rather than spread out across a wealth of species
responding independently to environmental change (42, 63). Further,
there is some evidence that maintaining evenness is important for
supporting multifunctionality in communities (64). These communities are
therefore of particular concern, especially considering evidence
evenness is more sensitive to environmental change that richness (62).

>>>>

\begin{abstract}
Aim: Despite unprecedented environmental change due to anthropogenic
pressure, recent work has found increasing dissimilarity due to turnover
but no overall trend in species diversity through time at the local
scale. Functional diversity provides a potentially powerful alternative
approach for understanding community composition by linking shifts in
species identity to mechanisms of ecosystem processes. Here we present
the first multi-taxa, multi-system analysis of functional change through
time.

@@

Results: We found no overall trend in any functional diversity metric,
which held even after correcting for changes in species richness. At the
study-level, there was substantial heterogeneity in the direction and
type of functional change exhibited. None of the included study
characteristics such as taxa, realm, biome, or protection status were
significantly related to the type of change exhibited by a community.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, the majority group of
studies showed no species or functional trends, but other types of
change were manifest for multiple studies, including studies
experiencing changes in functional redundancy, increased species and
functional richness, or loss of functional richness. With no one
prevailing scenario of change, it will be critical to link change
scenarios to ecological context.
\end{abstract}

@@

\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.}
Global analyses have been heavily criticized for geographic biases, lack
of data in the most heavily impacted areas, and exclusion of individual
studies' ecological context (10--12). Many of these criticisms reflect
limitations of ecological data on the whole,
\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
(13, 14).

@@

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While \textcolor{blue}{loss of functional diversity} is frequently
cited as one of the most pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
\textcolor{blue}{the functional space} shows mixed responses to
environmental change and extinction events (30, 31), with significant
impacts of species extinctions on functional diversity in some taxa and
not others (32). For some time periods, functional diversity appears to
be maintained for substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
(34).
\textcolor{blue}{Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.}

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as
\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
examining trends in plants, invertebrates, and other vertebrate species
is of equal interest, trait data for those taxa raise additional
challenges such as limited and biased species coverage (35), a lack of
accepted species-level means, and differences in the types of traits
collected. To ensure comparability across taxa in trait type and data
quality we therefore focus on mammals and birds.
\textcolor{blue}{We include body mass, foraging and dietary traits} that
were intentionally selected to be representative of a species' Eltonian
niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series
\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations. }
Time series represent full assemblages rather than populations of single
species (31), and all time series include abundance of observed species.
Following best practices for the database (37), studies with multiple
sample locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

@@

Our final dataset included 2,432 time series from 50 studies in 21
countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five biomes
(Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
earliest sample was in 1923 and the most recent was in 2014.
\textcolor{blue}{While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level. However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.}
For a full breakdown of studies and their characteristics, see the
supplement. Our final dataset reflects many of the data biases that make
global synthesis work challenging, including geographic bias, a bias
away from areas currently under the greatest threat, and a bias towards
shorter time series. We address these shortcomings and their potential
impact on our results in the discussion.

@@

We calculated yearly metrics of functional and species diversity for
each time series. Species-based metrics include species richness
(\emph{S}) and Jaccard similarity (\emph{J}) as a measure of turnover.
Jaccard similarity was calculated relative to the first observed year
for a time series. A negative trend in \emph{J} would therefore indicate
\textcolor{blue}{decreasing similiarity}. We did not impose a correction
for unobserved species as non-parametric estimators do not assign
species identities to corrected richness values, and therefore could not
be propagated to the functional diversity metrics.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (46). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species. `
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.
\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}

@@

We estimated general trends
\textcolor{blue}{across bird and mammals communities} for each diversity
metric using a linear mixed effects model with a random slope and
intercept for each study and each time series nested within the study,
\textcolor{blue}{methods which deal well with the inherent imbalances in our data.}
We fit 18 individual \emph{CWM} models,
\textcolor{blue}{one for each trait included in the analsysis}. All time
series with data for a given trait were included in the corresponding
\emph{CWM} model. We estimated study level trends using individual
linear models. For studies with more than one times series we fit a
random slope and intercept for time series. Some study-level models
could not be fit for five studies for at least one metric due to data
limitations, but those studies were still included in the general
models. They represented 12 of 1350 study-level models fit for each
metric. For further details see the supplement. Where appropriate,
response variables were \(log\) or \(log(x+1)\) transformed to better
fit model assumptions of residual normality.

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 47).
\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
We assessed the impact of time series duration and start year on
study-level trends using linear models with duration and start year as
predictors. All models in our analysis were fit using the \emph{lme4}
(v1.1-30) package in R (v4.2.3) and p-values were calculated by
Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(48--50).

@@

We found no significant overall \textcolor{blue}{temporal} trend in
species richness or functional diversity metrics
\textcolor{blue}{including functional richness, evenness, or divergence}
(observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall \textcolor{blue}{temporal}trends
indicate that although some studies experience increasing or decreasing
trends, the average trend across studies
\textcolor{blue}{was not significantly different from zero}` (Table
\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
different taxa, biomes, realms, or protection statuses were also
non-significant for richness and functional diversity metrics, with the
exception of a significantly increasing trend for functional evenness of
marine studies, and a significantly decreasing functional richness slope
for mammal studies. However, trends were not exhibited for the
\textcolor{blue}{corrected} metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. The general trends for \emph{CWM} models
were similarly not significant, with a significant positive trend for
only percentage of fish in diet composition.

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/3met_long} \caption{Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall mean slope for a metric in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES}\label{fig:timeseriesPlot}
\end{figure}

We did find significant differences between taxa, realms, and biomes for
Jaccard similarity and some of the \emph{CWM's}. For example, while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, Terrestrial,
Temperate, and Tropical communities, there was no significant slope for
the marine realm, indicating that the general trend is mostly driven by
turnover in terrestrial communities. We also found a significant
decrease in Jaccard similarity in unprotected areas only, with no trend
for protected areas. We found significant dietary shifts across
communities, with a significant general trend of increasing fish
consumption, which was also reflected in increasing fish consumption
trends for Bird studies and Temperate studies. We found a significantly
increasing trend in nectar consumption for Temperate studies but a
significantly decreasing tend for Tropical studies. We found
significantly decreasing trend for vertebrate consumption in Marine
studies and Tropical studies.

At the study level, 16 of 50 studies exhibited a significant trend in
species richness and 11 exhibited significant turnover. For uncorrected
functional diversity metrics, 25 of 50 studies exhibited a trend in a
least one metric, and 15 of 50 studies exhibited a significant trend for
at least one \textcolor{blue}{corrected} metric (Table
\ref{tab:trendTab}). In general, there were more significant trends for
uncorrected metrics, with some disappearing after correction, indicating
that those trends were likely due to changes in the number of species.
Hypothesis testing for study-level trends is likely affected by multiple
testing issues and some trends identified as significant are therefore
potentially spurious. Rather than interpreting changes in specific
studies, we present these results as a general picture of the kinds of
trends experienced by communities.

Study-level slopes were significantly related to start year of the time
series for Jaccard similarity and functional evenness models, both of
which had significantly more negative slopes with more recent start
year. No functional diversity metrics were significantly related to the
duration of the time series.

We assessed the sensitivity of general trend results to major data
processing decision by rerunning models with increasingly conservative
subsets of the data. Trends for Jaccard similarity and fish consumption
were not sensitive to either time series duration or trait coverage.
After excluding time series with less than three years we found an
increasing trend for body mass that remained after excluding time series
of less than four and five years. We also found a decreasing trend in
functional richness for models with a minimum of three years and a
minimum of five years. A decreasing functional richness trend was found
for models with a min of 85\% trait coverage, but not the other two
coverage levels. A complete list of models run in the sensitivity
analysis and their results can be found in the supplement.

@@

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
We did not detect an overall trend in any functional diversity metric,
\textcolor{blue}{corrected or uncorrected}. As with previous
species-based syntheses, we also found no overall trend in species
richness accompanied by \textcolor{blue}{increasing dissimilarity}
through time (31), indicating that non-significant trends in functional
metrics are consistent with similar well-documented species derived
trends. We found significant turnover for all biomes, realm, and taxa
except for marine studies, which stands in contrast to other global
estimates of biodiversity change that found higher turnover in marine
systems than terrestrial (37). However, previous global estimates are
dominated by fish communities which we exclude here and may be driving
the overall trend while disguising relative stasis in marine bird and
mammal communities.

We found evidence of functional richness loss in mammal studies, with a
loss trend significantly different from zero and significantly different
from the trend for birds, indicating a potential loss of functional
capacity in these communities. This result is consistent with previous
work linking anthropogenic drivers such as habitat loss, poaching, and
human modification and development to loss of functional diversity in
mammal communities across the globe (51--53). However we did not find
trends outside the random expectation for species loss, in contrast to
expectations for multiple continents based on IUCN risk categories (51).
In general, functional capacity of mammals shows greater declines that
bird communities.

We also found evidence of increasing evenness for Marine studies. The
empirical link between changes in evenness and ecosystem process is the
most poorly studied of the metrics, however theory indicates that
increased evenness improves function and stability (54). If one or a few
species become largely dominant in a community, function will be the
result of only those species rather than spread out across a wealth of
species responding independently to environmental change (42, 55).
Further, there is some evidence that maintaining evenness is important
for supporting multifunctionality in communities (56). The majority of
marine studies are of seabird communities (7 of 9), so the general trend
may be an early indicator of pay off from investments in sea bird
recovery, which is also indicated in recent species-based work (57).

Compositional shifts shown by significant trends in \emph{CWM'}s were
restricted to dietary traits, with no evidence of a trend in foraging
traits or body mass. Increases in fish consumption for bird and
temperate studies my be the product of increased availability of easy
fish from commercial fishing operations as fish are processed or
stocking of inland waterways for sport fishing. However, fisheries also
have negative impacts on waterbirds through competition for fish or bird
mortality due to by catch. Decreases in nectar consumption for tropical
studies is likely due declines in bat species, as 4 of the 5 tropical
studies are for mammal and bats were the only nectar consuming mammals
sampled in our data. Interpretation of the increases for temperate
studies is less clear, as temperate studies included many bird
communities with a variety of nectar consumers. Similarly, decreasing
trends for vertebrate consumption in marine and tropical studies are not
obviously consistent with known changes in those contexts and warrant
further examination.

@@

\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens. The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level. In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend. While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover. We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives. We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.}

@@

The lack of trend could be the result of multiple possible scenarios.
First, these may be communities resisting perturbations or simply not
experiencing significant perturbations. Given the studies in this group
come from all possible taxa, realms, biomes, and protection statuses,
evidence points to communities resisting perturbation, offering some
hope that there are areas of the globe where communities are fairing
well for now. Alternatively, these may be communities that have
experienced or continue to experience significant stress, but lost
species or functional diversity outside the observation window. This
could be true particularly for North American mammal communities where
trophic downgrading and megafaunal losses occurred hundreds of years ago
{[}citation{]}. Third, these communities may be experiencing directional
shifts undetectable by available data. For example, species-level trait
data does not capture intraspecific shifts in the trait space, which can
represent significant changes in the functional
\textcolor{blue}{space as a whole} and impact maintenance of ecological
processes.

Three of the groups fit under the broad umbrella of changes in
redundancy. By definition, if a community gains or loses species while
functional metrics are unchanged, those species represent an increase or
decrease in redundancy, so we include communities exhibiting gains or
losses in species richness with no change in functional metrics or
turnover and species richness loss with significant turnover in this
umbrella. For communities exhibiting loss of redundancy (declining
species richness), ecological processes are likely being maintained, but
capacity to respond to future stressors is reduced (58). These
communities are actually fairing better than expected looking at
species-based metrics alone, but are also in a precarious position for
maintaining ecological function into the future (59). Conversely,
communities exhibiting a increase in redundancy (via an increase in
species richness) are becoming better positioned to respond to future
stressors, but are not actually expanding their functional capacity as
may be assumed by looking at species gains alone (60).

The next group of studies exhibit an increase in species richness and
functional richness with significant turnover. While these communities
are losing some species that are functionally redundant or functional
analogs of species additions, they are gaining even more species that
expand the functional space. These communities have a hopeful
trajectory, as they are improving their functional capacity and
potentially their functional redundancy, likely leading to robust future
ecological function. Notably, the studies that exhibited species
richness increases (in this group with increasing functional richness,
as well as communities without functional richness increases) were
exclusively from terrestrial, temperate (sometimes temperate/polar) bird
communities and occurred in countries where there has been a significant
investment in conservation over the last few decades (United States,
Canada, Sweden). Our results are consistent with other functional work
in these regions, showing increases in species richness and functional
diversity for North American breeding birds over the last few decades
after a period of decline(61), and loss of common, functionally general
species even as rare species are increasing in North America and Europe
(62--64).

The final group exhibits decreasing functional evenness. These
communities are of particular concern not just because of the role
evenness plays in maintaining functionality, but because of increasing
evidence evenness is more sensitive to environmental change than species
richness (54). Formal tests of evenness changes as an early warning
signal of more catastrophic shifts in functional diversity will be
critical next step for understanding the full implication of evenness
loss for ecosystem processes.
